---
title: "DataCamp Data Manipulation in R"
author: "John Navarro"
date: "April 10, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#1. Intro to dplyr and tbl
```{r}
# Load the dplyr package
library(dplyr)

# Load the hflights package
library(hflights)

# Call both head() and summary() on hflightsh
head(hflights)
summary(hflights)
```
convert data.frame to tbl
```{r}
# Both the dplyr and hflights packages are loaded

# Convert data.frame hflights to tbl hflights
hflights<-tbl_df(hflights)

# Display head of tbl hflights

print(hflights)

# save column - UniqueCarrier as object named carriers
carriers<-hflights$UniqueCarrier
```
Changing labels of hflights part 1
```{r}
# Both the dplyr and hflights packages are loaded into workspace
lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")

# Add new column (Carrier) to hflights by combining lut with UniqueCarrier
hflights$Carrier<- lut[hflights$UniqueCarrier]

# Used glimpse()
glimpse(hflights)
```
changing labels of hflights part 2
```{r}
# examine codes in CancellationCode
unique(hflights$CancellationCode)

# the lookup table
lut <- c("A"="carrier", "B"="weather","C"="FFA","D"="security","E"="not cancelled")

# create a new vector Code, that contains the relabeled cancellation codes
hflights$Code <- lut[hflights$CancellationCode]

# check that column was added
glimpse(hflights)
```

# 2. select and mutate

Select(subsets by column), mutate(adds column), (variables)
filter(subsets by row), arrange(reorders)  (observations)
summarise( reduce by aggregating statistics) (groups)

```{r}
h1 <- filter(hflights, Carrier=="American")
#summarise(h1, mean)
```

The select verb
```{r}
# use select to print a tbl that only contains 4 columns
select(hflights, ActualElapsedTime, AirTime, ArrDelay, DepDelay)

# print a tbl with columns Origin up to and including Cancelled of hflights
select(hflights, 14:19)

# select columns YEar to Day of week, and ArrDelay to Diverted
select(hflights, 1:4,12:21)
```
Helper functions for variable selection
```{r}
# use contains helper function to print 2 columns
select(hflights, contains("Delay"))

# print out 5 columns
select(hflights,UniqueCarrier, ends_with("Num"), contains("Cancel"))

# print out 6 columns
select(hflights, ends_with("Time"), ends_with("Delay"))
```
Comparison to base R
```{r}
ex1r <- hflights[c("TaxiIn","TaxiOut","Distance")]
ex1d <- select(hflights, starts_with("Taxi"), Distance)

ex2r <- hflights[c("Year", "Month","DayOfWeek","DepTime","ArrTime")]
ex2d <- select(hflights, 1,2,4, ends_with("Time"))

ex3r <- hflights[c("TailNum","TaxiIn","TaxiOut")]
ex3d <- select(hflights, TailNum, starts_with("Taxi"))
```
2nd verb mutate
```{r}
# create a new df g1, with new col Actual GroundTime
g1 <- mutate(hflights, ActualGroundTime= ActualElapsedTime - AirTime)

# add another colunn groundTime
g2 <- mutate(g1, GroundTime= TaxiIn+TaxiOut)

# add another variable named AverageSpeed
g3 <- mutate(g2, Distance/AirTime *60)

# print
print(g3)

```
Add multiple variables using mutate
```{r}
# add column loss and loss_ratio to m1
m1 <- mutate(hflights, loss=ArrDelay - DepDelay, loss_ratio = loss/DepDelay)

# create m2 , add 3 variables
m2 <- mutate(hflights, TotalTaxi = TaxiIn +TaxiOut, ActualGroundTime = ActualElapsedTime - AirTime, Diff = TotalTaxi - ActualGroundTime)

unique(select(m2,Diff))
```
3rd verb Filter
```{r}
# print the column names of tbl
names(hflights)

# print out the flights that travelled a distance of 3k miles or more
filter(hflights, Distance>=3000 )

# print out the flights by 3 airlines
filter(hflights, Carrier == c("JetBlue","Southwest","Delta"))

# extract flights where taxiing took longer than flight
filter(hflights, (TaxiIn + TaxiOut)>AirTime)
```




























