---
title: "DataCamp Data Manipulation in R"
author: "John Navarro"
date: "April 10, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#1. Intro to dplyr and tbl
```{r}
# Load the dplyr package
library(dplyr)

# Load the hflights package
library(hflights)

# Call both head() and summary() on hflightsh
head(hflights)
summary(hflights)
```
convert data.frame to tbl
```{r}
# Both the dplyr and hflights packages are loaded

# Convert data.frame hflights to tbl hflights
hflights<-tbl_df(hflights)

# Display head of tbl hflights

print(hflights)

# save column - UniqueCarrier as object named carriers
carriers<-hflights$UniqueCarrier
```
Changing labels of hflights part 1
```{r}
# Both the dplyr and hflights packages are loaded into workspace
#lut <- c("AA" = "American", "AS" = "Alaska", "B6" = "JetBlue", "CO" = "Continental", 
         "DL" = "Delta", "OO" = "SkyWest", "UA" = "United", "US" = "US_Airways", 
         "WN" = "Southwest", "EV" = "Atlantic_Southeast", "F9" = "Frontier", 
         "FL" = "AirTran", "MQ" = "American_Eagle", "XE" = "ExpressJet", "YV" = "Mesa")

# Add new column (Carrier) to hflights by combining lut with UniqueCarrier
#hflights$Carrier<- lut[hflights$UniqueCarrier]

# Used glimpse()
glimpse(hflights)
```
changing labels of hflights part 2
```{r}
# examine codes in CancellationCode
unique(hflights$CancellationCode)

# the lookup table
#lut <- c("A"="carrier", "B"="weather","C"="FFA","D"="security","E"="not cancelled")

# create a new vector Code, that contains the relabeled cancellation codes
#hflights$Code <- lut[hflights$CancellationCode]

# check that column was added
glimpse(hflights)
```

# 2. select and mutate

Select(subsets by column), mutate(adds column), (variables)
filter(subsets by row), arrange(reorders)  (observations)
summarise( reduce by aggregating statistics) (groups)

```{r}
h1 <- filter(hflights, Carrier=="American")
#summarise(h1, mean)
```

The select verb
```{r}
# use select to print a tbl that only contains 4 columns
select(hflights, ActualElapsedTime, AirTime, ArrDelay, DepDelay)

# print a tbl with columns Origin up to and including Cancelled of hflights
select(hflights, 14:19)

# select columns YEar to Day of week, and ArrDelay to Diverted
select(hflights, 1:4,12:21)
```
Helper functions for variable selection
```{r}
# use contains helper function to print 2 columns
select(hflights, contains("Delay"))

# print out 5 columns
select(hflights,UniqueCarrier, ends_with("Num"), contains("Cancel"))

# print out 6 columns
select(hflights, ends_with("Time"), ends_with("Delay"))
```
Comparison to base R
```{r}
ex1r <- hflights[c("TaxiIn","TaxiOut","Distance")]
ex1d <- select(hflights, starts_with("Taxi"), Distance)

ex2r <- hflights[c("Year", "Month","DayOfWeek","DepTime","ArrTime")]
ex2d <- select(hflights, 1,2,4, ends_with("Time"))

ex3r <- hflights[c("TailNum","TaxiIn","TaxiOut")]
ex3d <- select(hflights, TailNum, starts_with("Taxi"))
```
2nd verb mutate
```{r}
# create a new df g1, with new col Actual GroundTime
g1 <- mutate(hflights, ActualGroundTime= ActualElapsedTime - AirTime)

# add another colunn groundTime
g2 <- mutate(g1, GroundTime= TaxiIn+TaxiOut)

# add another variable named AverageSpeed
g3 <- mutate(g2, Distance/AirTime *60)

# print
print(g3)

```
Add multiple variables using mutate
```{r}
# add column loss and loss_ratio to m1
m1 <- mutate(hflights, loss=ArrDelay - DepDelay, loss_ratio = loss/DepDelay)

# create m2 , add 3 variables
m2 <- mutate(hflights, TotalTaxi = TaxiIn +TaxiOut, ActualGroundTime = ActualElapsedTime - AirTime, Diff = TotalTaxi - ActualGroundTime)

unique(select(m2,Diff))
```
#3. 3rd verb Filter
```{r}
# print the column names of tbl
names(hflights)

# print out the flights that travelled a distance of 3k miles or more
filter(hflights, Distance>=3000 )

# print out the flights by 3 airlines
filter(hflights, Carrier == c("JetBlue","Southwest","Delta"))

# extract flights where taxiing took longer than flight
filter(hflights, (TaxiIn + TaxiOut)>AirTime)
```
Combining tests using Boolean operators
```{r}
# select the rows where a flight left before 5am or after 10p
filter(hflights, DepTime < 500 | ArrTime >2200)

# Print out all flights that departed late but arrived ahead of scheduel
filter(hflights, DepDelay >0 & ArrDelay<0)

# find all the flights that were cancelled after being delayed
filter(hflights, Cancelled ==1 & DepDelay>0)
```
Combine use of first three verbs
```{r}
# First use filter to select flights that had JFK as destination
c1 <- filter(hflights, Dest == "JFK")

# Add a new column named date, 
c2 <- mutate(c1, Date = paste(Year, Month, DayofMonth, sep='-'))

# select columns to provide an overview
select(c2, Date, DepTime, ArrTime, TailNum)

```

Fourth verb, arrange, arranges by values in a column
```{r}
# define dtc
dtc <- filter(hflights, Cancelled == 1, !is.na(DepDelay))

# arrange dtc by departure delays
arrange(dtc, DepDelay)

# arrange so that cancellation codes are grouped
arrange(dtc, CancellationCode)

# arrange by carrier, tiebreaker is dep delay
arrange(dtc, Carrier, DepDelay)
```
Reversing the order of arranging
```{r}
# arrange hflights by carrier and reverse dep delay
arrange(hflights, Carrier, desc(DepDelay))

# arrange flights in hflights by their total delay
arrange(hflights, DepDelay + ArrDelay)
```
Recap on the first 4 verbs

```{r}
# print only the Tail Num of hflights for flights that departed too late and sorted by total taxi time
c1 <- filter(hflights, DepDelay>0)
c2 <- mutate(c1, TotalTaxi = TaxiIn+TaxiOut)
c3 <- arrange(c2, TotalTaxi)
select(c3, TailNum)

# In one line
select(arrange(mutate(filter(hflights, DepDelay>0), TotalTaxi = TaxiIn + TaxiOut), TotalTaxi), TailNum)
```

#4. 5th verb Summarise - does NOT return an altered copy of the tbl
```{r}
# use summarise, min dist and max dist
summarise(hflights, min_dist = min(Distance), max_dist = max(Distance))

# print out a summary with variable max_div
summarise(filter(hflights, Diverted ==1), max_div = max(Distance))
```
Aggregate functiosn
```{r}
# Remove ros that have NAs in the arrival delay column save to temp1
temp1 <- filter(hflights, !is.na(ArrDelay))

# print out a summary of temp1
summarise(temp1, earliest = min(ArrDelay), average = mean(ArrDelay),
          latest = max(ArrDelay), sd = sd(ArrDelay))

# filter out NA in TaxiInand TaxiOUt, save to temp2
temp2 <- filter(temp1, !is.na(TaxiIn) & !is.na(TaxiOut))

# Print out a summary of temp2
summarise(temp2, max_taxi_diff = max(abs(TaxiIn - TaxiOut)))
```

dplyr aggregate functions
```{r}
# print out a summary of hflights
summarise(hflights, n_obs= n(), n_carrier=n_distinct(Carrier), n_dest=n_distinct(Dest))

# create aa
aa <- filter(hflights, Carrier == "American")

# print out a summary of aa
summarise(aa, n_flights=n(), n_canc=sum(Cancelled ==1), avg_delay=mean(ArrDelay, na.rm=TRUE))
```

Using pipe operator to string together commands

```{r}
# write a piped version
hflights%>%mutate(diff = TaxiOut - TaxiIn)%>%filter(!is.na(diff))%>%summarise(avg=mean(diff))
```

Drive of Fly? Part 1 of 2
```{r}
# use pipe to chaing together dplyr verbs to compare avg air velocity to avg car velocity
hflights%>%mutate(RealTime=100+ActualElapsedTime, mph=Distance/RealTime * 60)%>%
  filter(!is.na(mph) & mph<70)%>%
  summarise(n_less=n(), n_dest=n_distinct(Dest), min_dist=min(Distance), max_dist=max(Distance))
# These are the flights that travel slowly (less than 70mph)
```
Drive or fly part 2
```{r}
# Determine which flights are slower than 105 mph
hflights%>%mutate(RealTime=100+ActualElapsedTime, mph=Distance/RealTime * 60)%>%
  filter(mph<105 |Cancelled==1|Diverted==1)%>%
  summarise(n_non = n(),
            n_dest=n_distinct(Dest),
            min_dist=min(Distance),
            max_dist=max(Distance))
```
Another piping exercise - How many flights were overnight flights
```{r}
hflights%>%filter(!is.na(DepTime), !is.na(ArrTime), DepTime > ArrTime)%>%
  summarise(num=n())
```
# 5. Using Group by 

Use group by and summarise together. Automatically adds column containing variable grouped by

```{r}
# Make an ordered per-carrier summary of hflights
hflights%>%
  group_by(UniqueCarrier)%>%
  summarise(p_canc= mean(Cancelled==1) * 100, avg_delay=mean(ArrDelay, na.rm=TRUE))%>%
  arrange(avg_delay, p_canc)
```
Combine group by with mutate

```{r}

```






















